---
title: "Race and Income in California"
author: "Sebastian Ramirez Feune"
description: "An examination of inequality in California"
output:
  distill::distill_article:
    self_contained: false
---
```{r include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE, 
  fig.align = "center"
)
```


## Introduction

Despite its diverse population, California continues to be a center for significant social and economic inequality. Why is this so? It is common to look back to a history of oppression and discrimination that created institutions that continue to play integral roles in American society. This study aims to closely examine the relationship between race and income. How is income level generally related to race? And, what are some factors that may be impacting these income disparities? Thus, in addition to observing how income generally correlates with race, I will be simultaneously tracking how race is correlated with education level and unemployment. 

I contend that there were will be strong correlations made between income and racial identity in California, much like the rest of the United States. Black, Native American, and Hispanic/Latino communities will likely be in the lower ranks of income level, along with education and employment levels. This is largely due to historic discrimination of these two communities resulting in institutional biases and racism barring these demographics from many resources available to their Asian and white counterparts. While Asian Americans are people of color and recipients of institutionalized racism, their designation as “model minorities” and general arrival to the United States with high levels of education has aided in their ascension into higher income brackets. This is generally interesting because it gives concrete evidence and examples as to how the overall lack of access to resources has a strong link with the income inequality that we see in our nation.


## Data 

In this cross-sectional study, we are using two sources: 

1. The American Community Survey (ACS) of 2020 was used to attain  data on income, education, and unemployment

2. The Decennial Census of 2020 was used to attain data on 

Both sources were merged together into the same data set.

I used four variables: race, income, education, and unemployment. The state that we will be specifically observing is California, narrowing the scope of the study. The key independent variable is race: Black, white, Hispanic, Asian, and Native American. Using these variables, we are able to measure the dependent variables, which include income, education, and unemployment. 

```{r}

library(tidyverse)
library(tidycensus)
library(sf)
library(ggplot2)

census_api_key('5e4ef03bc94c689219c200d6b833aa56803bb648')
```



```{r}
options(tigris_use_cache = TRUE)

ca_income <- get_acs(
  geography = "tract", 
  variables = c(income = "B19013_001",
  bachelor = "B06009_005", 
  unemployment = "B23025_005"),
  state = "CA", 
  year = 2020,
  geometry = TRUE
)


california_race <- get_decennial(
  geography = "tract", 
  state = "CA", 
  variables = c(
    Hispanic = "P2_002N",
    White = "P2_005N",
    Black = "P2_006N",
    Native = "P2_007N",
    Asian = "P2_008N",
    Total = "P2_001N"
  ),
  year = 2020,
  geometry = TRUE
)

```

```{r}
options(tigris_use_cache = TRUE)

ca_income1 <- get_acs(
  geography = "county", 
  variables = "B19013_001",
  state = "CA", 
  year = 2020,
  geometry = TRUE
)


```

```{r}
ca_income <- ca_income |>
  select(-moe) |>
  pivot_wider(
    names_from = "variable", 
    values_from = "estimate"
  )
california_race <- california_race |>
  pivot_wider(
    names_from = "variable", 
    values_from = "value")

cali_merge <- merge(x = as.data.frame(california_race), y = as.data.frame(ca_income), by.x = "GEOID", by.y = "GEOID")

cali_merge <- cali_merge |>
  mutate(percentage_white = (White/Total)*100, percentage_black = (Black/Total)*100, percentage_hispanic = (Hispanic/Total)*100, percentage_asian = (Asian/Total)*100, percentage_native = (Native/Total)*100, percentage_unemployed = (unemployment/Total)*100, percentage_bachelor = (bachelor/Total)*100)

```

```{r}
plot(ca_income["income"], main = "Average Income by California Tract")

plot(ca_income1["estimate"], main = "Average Income by California County")
```
```{r}
cali_race <- get_decennial(
  geography = "county", 
  state = "CA", 
  variables = c(
    Hispanic = "P2_002N",
    White = "P2_005N",
    Black = "P2_006N",
    Native = "P2_007N",
    Asian = "P2_008N"
  ),
  summary_var = "P2_001N",
  year = 2020,
  geometry = TRUE
) %>%
  mutate(percent = 100 * (value / summary_value))


```

```{r}
cali_race |>
  ggplot() +  
  geom_sf(aes(fill= percent)) +  
  scale_fill_viridis_c(option = "H") +  
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), 
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  facet_wrap(~variable, ncol = 2)
```


Above are three visuals of the primary dependent and independent variables: race and income. The first two maps illustrate the distribution of incomes throughout the state of California. While the study observes income distribution based on census tracts, one map shows distribution across counties as a simpler version of the census tract-based model. These models reveal how wealth tends to be concentrated in coastal regions, more specifically the San Francisco Bay Area.

The third map shows California's racial makeup throughout California counties (for simplicity sake). The race map shows the average racial makeup of each California tract/county, with Hispanic Californians spanning across California, especially near the coast, Southern California, and Central Valley. The second most prominent ethnic/racial group is white, with predominantly white tracts concentrating in more Eastern and Northern areas of the state. Asian Americans tend to be found near Los Angeles and the San Francisco Bay Area. Finally, the Black and Native Americans are relatively small throughout the state.

## Results

```{r, results = 'asis', echo = FALSE, eval=TRUE, warning = FALSE, message = FALSE}
##library(stargazer)

m.lm_black <-  lm(income ~  percentage_black + percentage_bachelor +  percentage_unemployed, data = cali_merge)

m.lm_white <- lm(income ~ percentage_bachelor +  percentage_unemployed + percentage_white, data = cali_merge)

m.lm_hisp <-  lm(income ~  percentage_hispanic + percentage_bachelor +  percentage_unemployed, data = cali_merge)

m.lm_asian <-  lm(income ~  percentage_asian + percentage_bachelor +  percentage_unemployed, data = cali_merge)

m.lm_native <-  lm(income ~  percentage_native + percentage_bachelor +  percentage_unemployed, data = cali_merge)

##stargazer(m.lm, type = "html")
library(modelsummary)
cm <- c("percentage_bachelor" = "Percentage Bachelor's", "percentage_unemployed" = "Percentage Unemployed", "percentage_black" = "Percentage Black", "percentage_white" = "Percentage White", "percentage_hispanic" = "Percentage Hispanic", "percentage_asian" = "Percentage Asian", "percentage_native" = "Percentage Native")
modelsummary(list("Tract Income 1" = m.lm_black, "Tract Income 2" =  m.lm_white, "Tract Income 3" =  m.lm_hisp, "Tract Income 4" =  m.lm_asian, "Tract Income 5" =  m.lm_native), coef_map = cm, stars = TRUE)
```

This table illustrates the regressions of multiple variables in relation to income. All of the regression show strong correlations and significance, as indicated by the three asterisks adjacent to every value. This indicates that there is likely a causal link between the values. When it comes to income, the linear regressions reveal that with a single percentage increase of an Asian or white person in a tract will increase the average Asian and white incomes, respectively, more than those of the Black, Hispanic, and Native populations. Meanwhile, the average income of Hispanic, Black, and Native American (in order of increasing income loss) decreases with each additional single percentage increase in their respective populations.

More specifically, for every single percentage increase in the white population, the income of the white population increases by \$675.437. For every single percentage increase in the Black population, the income of the Black population decreases by \$1242.166 For every single percentage increase in the Hispanic population, the income of the Hispanic population decreases by \$887.981. For every single percentage increase in the Asian population, the income of the Asian population increases by \$917.810. For every single percentage decrease in the Native population, the income of the Native population increases by \$3196.215. Then, when looking at Bachelor's Degrees, the table shows that for every single percent increase in the portion of residents holding a Bachelor's in the Census tract increases tract income by 273.868. Meanwhile, increasing the unemployment rate by 1% decreases average income by \$81.783.


```{r}
cali_plot <- ggplot(cali_merge, aes(y = Total)) + 
  geom_smooth(method = "lm", aes(x = percentage_white ), color = "orange", fill = "orange", alpha = 0.25) + 
  geom_smooth(method = "lm",aes(x = percentage_black), color = "purple", fill = "purple", alpha = 0.25) + 
  geom_smooth(method = "lm",aes(x =percentage_asian ), color = "turquoise", fill = "turquoise", alpha = 0.25) + 
  geom_smooth(method = "lm",aes(x = percentage_native), color = "blue", fill = "blue", alpha = 0.25)+
 geom_smooth(method = "lm",aes(x =percentage_hispanic ), color = "red", fill = "red", alpha = 0.25) +
   labs(x = "Percentage of Each Race (Tract)", y = "Income (dollars)", title = "Tract Income across Racial Lines", caption = "Orange = white | Purple = Black | Turquoise = Asian | Blue = Native | Red = Hispanic") 
cali_plot 
```


This graph illustrates the average income levels for Asian, Black, White, Hispanic, and Native residents of California according to the 2020 Census. The graph shows interesting implications of both economic inequity and racial composition in the state of California. The graph indicates that both the Indigenous and Black populations are relatively low throughout California, with the Black population never exceeding 20% and the Indigenous population never exceeding 2% in any Census tract. 

The graph indicates that in low income regions where income is less than \$100,000 dollars, these areas tend to be predominantly Hispanic American, especially around the \$45,000 annual income mark. Asian, Black, and white Americans tend to be the same population proportion at an income of zero, but they begin to diverge from here forward. Consistently, tracts where Asian and white Californians are more of a majority correspond with increasingly higher wages. Hence, the more white and Asian an area is, the more likely those whites and Asians are going to be affluent. On the other hand, in places were Black and Hispanic Californians make more money, especially after \$150,000, they are likely to make up a smaller share of the population, at about 5% or less for Black Californians and about 15% or less for Latino Californians.


## Conclusion

Our analysis has shown that race and income in California are both strongly correlated. When there is an increase in the population of Black, Hispanic, or Indigenous people in a Census tract, the average income level of the tract is expected to drop. Meanwhile, an increase in Asian and white Californians increase the average income. Race and income are inextricably tied, and they reflect a much broader American issue of racial discrimination. 

While observing these results, I understood the importance of measuring other predictor variables, such as education level and unemployment levels of a certain tract. The positive correlation between income and attainment of a Bachelor's degree indicates how education level is a potential factor that may largely influence the dichotomies in income that the various racial groups face. With unemployment levels holding a negative relationship with income, it is clear that higher levels of unemployment in certain communities may effect significant income losses.

Examining these separate variables is essential because there is no inherent reason for which certain demographics trend towards lower affluence in a state like California. In addition to be racial/ethnic minorities, there are also institutional obstacles that many families of color face. Asian American families hold a peculiar spot within this demographic, for they do face institutional racism, but in a different manner, with many white Americans using Asian Americans as "model minorities." 

My intention with creating an appendix was to dive deeper into the relationship between race and both education and unemployment. While my main analysis observes income as the outcome variable, this appendixed analysis uses both education and unemployment as outcome variables. 




